DATA 621---Business Analytics and Data Mining

### 2020-09-14 HW1/HW1_draft_kellogg.Rmd
part 1 complete

---
title: 'DATA 621---Business Analytics and Data Mining'
subtitle: 'Fall 2020---Group 2---Homework #1'
author: Avraham Adler, Samantha Deokinanan, Amber Ferger, John Kellogg,
    Bryan Persaud, Jeff Shamp
date: "9/27/2020"
output:
  html_document: default
  pdf_document: default
---

@@

```{r loadData, include=FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(ggplot2)
library(corrplot)
library(data.table)
library(tidymodels)
library(mice)
library(psych)
train_set <- read.csv('data/moneyball-training-data.csv')
eval_set <- read.csv('data/moneyball-evaluation-data.csv')

@@

## 1. DATA EXPLORATION

@@

### What's missing? 
A first look at the data shows that only about 8% of the records have a full set
of information. The good news is that most of the missing values come from 
statistics that don't happen too often: hit-by-pitch (92% missing!), caught 
stealing (34% missing), and double plays (13% missing). Since we have so little 
hit-by-pitch data, we expect that it doesn't contribute much to overall wins and will 
eliminate it from a few of the models we propose. The other two stats have less 
than half of the data missing, so we'll need to think of a clever way to fill in 
these values. The remaining missing information is from a combination of stolen 
bases and strikeouts (by batters and pitchers). **It seems completely unreasonable** 
to have zero strike outs in a season, so this is something we'll most certainly have 
to impute.

@@

When we take a closer look at the data, these negative relationships aren't obvious. 
In fact, only Errors and Hits allowed seem to have a negative impact on wins. Caught 
stealing and Strikeouts by batters appear to be random; this means that whether the 
stat for a particular season is high or low doesn't affect the overall number of wins.

Even more interestingly, Home Runs allowed and Walks allowed have the *opposite* 
effect; as these stats increase, so do the number of wins!

@@

Many of these stats *do* seem to have an effect on the number of wins, most notably,
base hits and walks. We see weaker positive relationships for home runs, doubles, triples, 
and stolen bases. This makes sense when we think about it; these things tend to happen 
less often in games than pure base hits and walks, so they don't have as much of an 
effect on winning. Finally, double plays and batters hit don't appear to have any 
correlation with the number of wins. Once again, this intuitively makes sense because 
they are less likely to happen in a game.

@@

Before using any of the statistics in a model, we need to take a closer look at the 
variation in the data. We call out of the ordinary values (exceptionally high or
low values) **outliers**. We need to take these into account in our modeling because
we want to make sure our predictions aren't skewed because of them.

```{r summaryInfo, include = FALSE}

summary(train_set)
```

Some of the provided statistics are well-balanced in the sense that there are
very few (or no) extreme values. **Home Runs by batters**, **Strikeouts by** 
**batters**, and **Home Runs allowed** are examples of this.

```{r fewOutliers, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 2.5, fig.width = 5, fig.align = 'center'}

@@

Conversely, some of the stats have a very high number of outliers, indicating that there
are some seasons with some abnormally high or low values. **Errors**, **Hits allowed**,
**Stolen bases**, and **Walks by batters** are examples of this.

```{r manyOutliers, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 2.5, fig.width = 6, fig.align = 'center'}

@@

#sapply(train_set %>% select(-INDEX, -TARGET_WINS), function(x){length(boxplot.stats(x)$out)})

@@

* All of the outliers for **Errors** and **Hits allowed** are above the upper 
tail of the data set. This is further illustrated by the mean and median values
for both of these stats; in both instances, the mean per year (Errors = ~246/year
and Hits allowed = ~1779/year) are higher than the median per year (Errors = 
~159/year and Hits allowed = ~1518/year). This means that some seasons with
exceptionally high values for both of these statistics skew the dataset. 
* The maximum value

### Are stats correlated?
```{r, echo = FALSE, message = FALSE, warning = FALSE}

# Correlation matrix with significance levels (p-value)
library("Hmisc")
res2 <- rcorr(as.matrix(train_set %>% select(-INDEX)))

@@

## Data Preparation

Now we have a good idea of the data we are looking at we can take the next steps to 
prepare it for building a solid model.  We saw from the above that at least one 
variable should be omitted for low reporting of data. There are also outlier concerns 
for several of the variables, namely Team Pitching Strikeouts (`TEAM_PITCHING_SO`), 
Team Batting Hits (`TEAM_BATTING_H`) and Team Pitching Hits ('TEAM_PITCHING_H').

@@

## "MICE" imputation method and Feature Engineering

@@

## Modeling

We will start with a simple linear model and select out features based on null hypothesis \
testing for non-zero slope.

@@

>>>>

---
title: 'DATA 621---Business Analytics and Data Mining'
subtitle: 'Fall 2020---Group 2---Homework #1'
author: Avraham Adler, Samantha Deokinanan, Amber Ferger, John Kellogg,
    Bryan Persaud, Jeff Shamp
date: "9/27/2020"
output:
  pdf_document: default
  html_document: default
---

@@

```{r loadData, include=FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(ggplot2)
library(corrplot)
library(data.table)
library(tidymodels)
library(mice)
library(psych)
library(Hmisc)
train_set <- read.csv('https://raw.githubusercontent.com/aadler/DT621_Fall2020_Group2/master/HW1/data/moneyball-training-data.csv')
eval_set <- read.csv('https://raw.githubusercontent.com/aadler/DT621_Fall2020_Group2/master/HW1/data/moneyball-evaluation-data.csv')

@@

## DATA EXPLORATION

@@

### What's missing? 
A first look at the data shows that only about 8% of the records have a full set of 
information. The good news is that most of the missing values come from statistics 
that don't happen too often: hit-by-pitch (`TEAM_BATTING_HBP`, 92% missing!), caught 
stealing (`TEAM_BASERUN_CS`, 34% missing), and double plays (`TEAM_FIELDING_DP`, 13% 
missing). Since we have so little hit-by-pitch data, we expect that it doesn't contribute 
much to overall wins and will eliminate it from a few of the models we propose. The 
other two stats have less than half of the data missing, so we'll need to think of a 
clever way to fill in these values. The remaining missing information is from a 
combination of stolen bases and strikeouts by pitchers and batters (`TEAM_BASERUN_SB`,
`TEAM_PITCHING_SO`, `TEAM_BATTING_SO`). **It seems completely unreasonable** to have zero 
strike outs in a season, so this is something we'll most certainly have to impute.

@@

When we take a closer look at the data, these negative relationships aren't obvious. In 
fact, only **Errors** and **Hits allowed** seem to have a negative impact on wins. **Caught**
**stealing** and **Strikeouts by batters** appear to be random; this means that whether the 
stat for a particular season is high or low doesn't affect the overall number of wins.

Even more interestingly, **Home Runs allowed** and **Walks allowed** have the *opposite* 
effect; as these stats increase, so do the number of wins!

@@

Many of these stats *do* seem to have an effect on the number of wins, most notably,
**Base hits** and **Walks**. We see weaker positive relationships for **Home runs**, 
**Doubles**, **Triples**, and **Stolen bases**. This makes sense when we think about it; 
these things tend to happen less often in games than pure base hits and walks, so they don't 
have as much of an effect on winning. Finally, **Double plays** and **Batters hit** don't 
appear to have any correlation with the number of wins. Once again, this intuitively makes 
sense because they are less likely to happen in a game.

@@

Before using any of the statistics in a model, we need to take a closer look at the variation 
in the data. We call out-of-the-ordinary values (exceptionally high or low values) **outliers**. 
We need to take these into account in our modeling because we want to make sure our predictions 
aren't skewed because of them.

```{r summaryInfo, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 3, fig.width = 5, fig.align = 'center'}

#summary(train_set)

totalOutliers <- data.frame(sapply(train_set %>% select(-INDEX, -TARGET_WINS, - TEAM_BATTING_HBP), 
       function(x){length(boxplot.stats(x)$out)/nrow(train_set)}))

totalOutliers$VARIABLE_NM <- rownames(totalOutliers)
colnames(totalOutliers) <- c('PCT_OUTLIERS', 'VARIABLE_NM')

ggplot(totalOutliers , aes(x=reorder(VARIABLE_NM, PCT_OUTLIERS), y=PCT_OUTLIERS, label = round(PCT_OUTLIERS, 3))) + 
  geom_text(vjust = 0.5, hjust = -0.05)+
  geom_bar(stat = "identity") +
  ggtitle("Percent Outliers") +
  xlab("Statistic") + ylab("Percent of Data that is an Outlier") + 
  coord_flip() +
  expand_limits(y = 0.15)

As we can see, some of the provided statistics are well-balanced in the sense that there are very 
*few* (or no) extreme values. **Home Runs allowed** (`TEAM_PITCHING_HR`), **Strikeouts by batters** 
(`TEAM_BATTING_SO`), and **Home Runs by batters** (`TEAM_BATTING_HR`) are examples of this.

```{r fewOutliers, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 2, fig.width = 6, fig.align = 'center'}

@@

Conversely, some of the stats have a very *high* number of outliers, indicating that there
are some seasons with some abnormally high or low values. **Errors** (`TEAM_FIELDING_E`), 
**Hits allowed** (`TEAM_PITCHING_H`), **Walks by batters** (`TEAM_BATTING_BB`), and 
**Stolen bases** (`TEAM_BASERUN_SB`) are examples of this.

```{r manyOutliers, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 2, fig.width = 6, fig.align = 'center'}

@@

@@

* All of the outliers for **Errors**, **Hits allowed**, and **Stolen bases** are above the upper 
tail of the data set. This is further illustrated by the mean and median values for these stats; in 
all instances, the mean per year (Errors = ~246/year, Hits allowed = ~1779/year, Stolen bases = ~125/year) 
are higher than the median per year (Errors = ~159/year, Hits allowed = ~1518/year, Stolen bases = 
~101/year). This means that some seasons with exceptionally high values skew the dataset. 
* There are a few *very* extreme outliers for **Hits allowed**. The maximum value is 30,132, which
is over 16 times the average number of hits allowed per season!
* There are outliers both above *and* below the tails of the data for the **Walks by batters** stat. 
This means that we have exceptionally low (min = 0!) and exceptionally high (max = 878) seasons.

The remaining stats, **Walks allowed** (`TEAM_PITCHING_BB`), **Base Hits by batters** (`TEAM_BATTING_H`), 
**Caught stealing** (`TEAM_BASERUN_CS`), **Strikeouts by pitchers** (`TEAM_PITCHING_SO`), **Double plays** 
(`TEAM_FIELDING_DP`), **Triples by batters** (`TEAM_BATTING_3B`), and **Doubles by batters** 
(`TEAM_BATTING_2B`) have between 29 and 99 outliers.

```{r otherOutliers, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 3.5, fig.width = 7, fig.align = 'center'}

set3 <-  train_set %>%
  select('Base Hits by batters' = TEAM_BATTING_H, 
         'Doubles by batters' = TEAM_BATTING_2B,  
         'Triples by batters' = TEAM_BATTING_3B,
         'Caught stealing' = TEAM_BASERUN_CS,
         'Walks allowed' = TEAM_PITCHING_BB,
         'Strikeouts by pitchers' = TEAM_PITCHING_SO,
         'Double plays' = TEAM_FIELDING_DP) %>%
  gather("stat", "value") %>%
  filter(complete.cases(.) == TRUE)

vals3 <- ggplot(set3 , aes(x=stat, y=value)) + 
    geom_boxplot() +
  labs(title="",x="", y = "Value") +
  facet_wrap(~ stat, scale = 'free', ncol = 4)

vals3

Some things to note:

* All variables are very narrowly distributed, meaning that most of the data falls within
a small range. 
* **Strikeouts by pitchers** and **Walks allowed** have a few very extreme outliers; these 
represent seasons that have abnormally high values for the statistics. 
* The average number of pure **Base Hits** (1469/season) is greater than the average number
of Doubles (~241/season) and Triples (~55/season). This isn't at all surprising, but serves 
as a good gut check on the validity of the data.

### Are stats correlated?

We would expect that a few things in the dataset might be correlated: perhaps number of errors and
hits/homeruns allowed or the number of base hits by batters and doubles/triples/homeruns. We can
visualize the correlations between the statistics to determine if there is a significant relationship
between the them: blue dots represent positively correlated variables (as one increases,
so does the other) and red dots represent negatively correlated variables (as one increases,
the other decreases).

```{r, echo = FALSE, message = FALSE, warning = FALSE}

# Correlation matrix with significance levels (p-value)
res2 <- rcorr(as.matrix(train_set %>% select(-INDEX)))

@@

Some noteworthy relationships (coincidental or not):

* **Errors** are highly, negatively correlated with walks by batters, strikeouts by batters, and
homeruns (both by batters and allowed). 
* **Triples by batters** are highly, negatively correlated with strikeouts by batters and homeruns
(both by batters and allowed).
* **Strikeouts by batters** are highly, positively correlated with homeruns (both by batters and
allowed). 
* **Homeruns by batters** and **Homeruns allowed** are both positively correlated with walks by 
batters.
* As expected, **basehits** are positively correlated with doubles and triples.

We can keep these correlations in mind when developing our models: if we have correlated statistics, 
there could be in-built redundancy in the features, and we may be able to create a simpler, more
accurate model by eliminating some.

## DATA PREPARATION

Now we have a good idea of the data we are looking at we can take the next steps to 
prepare it for building a solid model.  We saw from the above that at least one 
variable should be omitted for low reporting of data. There are also outlier concerns 
for several of the variables, namely Team Pitching Strikeouts (`TEAM_PITCHING_SO`), 
Team Batting Hits (`TEAM_BATTING_H`) and Team Pitching Hits (`TEAM_PITCHING_H`).

@@

### "MICE" imputation method and Feature Engineering

@@

## BUILD MODELS

### Base Model
We will start with a simple linear model and select out features based on null hypothesis \
testing for non-zero slope.

@@

## SELECT MODELS
